A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-15 02-ceu.tex
Compress captions of figs and listings in Section 2

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1#2{\hfill\rmfamily\itshape\makebox[#1\columnwidth][l]{//~#2}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{1.}{Declarations}|
input $\T<type>$ $\T<id>$; | \C{.6}{declares an external input event} |
event $\T<type>$ $\T<id>$; | \C{.6}{declares an internal event}       |
var   $\T<type>$ $\T<id>$; | \C{.6}{declares a variable}              |

|\C{1.}{Event handling}|
$\T<id>$ = await $\T<id>$;   | \C{.6}{awaits an event and assigns the received value} |
emit $\T<id>$($\T<exp>$);    | \C{.6}{emits an event passing a value} |

|\C{1.}{Control flow}|
$\T<stmt>$ ; $\T<stmt>$                          | \C{.45}{sequence}     |
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end    | \C{.45}{conditional}  |
loop do $\T<stmts>$ end                      | \C{.45}{repetition}   |
every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end         | \C{.45}{event iteration}   |
finalize [$\T<stmts>$] with $\T<stmts>$ end       | \C{.45}{finalization} |

|\C{1.}{Logical parallelism}|
par/or  do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{aborts when any side ends}      |
par/and do $\T<stmts>$ with $\T<stmts>$ end | \C{.45}{terminates when all sides ends} |

|\C{1.}{Assignment \& Integration with C}|
$\T<id>$ = $\T<exp>$; | \C{.6}{assigns a value to a variable}              |
_$\T<id>$($\T<exps>$) | \C{.6}{calls a C function (id starts with `\_'\,)} |
\end{lstlisting}
\egroup

Listing~\ref{lst.ceu} shows a complete example in \CEU that toggles a LED
whenever a radio packet is received, terminating with a button press always
with the LED off.
%
The implementation first declares the \code{BUTTON} and \code{RADIO\_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single-statement trail that waits for a button press before terminating
(ln.~4), and an endless loop that toggles the LED on and off on radio receives
(ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off (ln.~7).

@@

\begin{figure}[b]
\centering
\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}

@@

@@

\begin{figure}[h]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=right]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are never concurrent.}
\end{minipage}%
%
\begin{minipage}[h]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=2em]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are concurrent but still deterministic.}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}%
}
\end{figure}

@@

\begin{figure}[h]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right]
par/or do
   var _msg_t msg;
   <...> // prepare msg
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
//
\end{lstlisting}
\small{\ax\!Local resource finalization}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=2em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{\bx External resource finalization}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for
local resources and \emph{memory leaks} for external resources.%
\label{lst.fin.ceu}}
\end{figure}

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
    \item If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
    \item If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

>>>>

\bgroup
\def\T<#1>{\langle\mathit{#1}\rangle}
\def\C#1{\rmfamily\itshape//~#1}
\def\CC#1#2{\hfill\makebox[#1][l]{\C{#2}}}
\begin{lstlisting}[
  basicstyle=\ttfamily\footnotesize,
  caption={The concrete syntax of \CEU.},
  label={lst.syntax},
]
|\C{Declarations}|
input $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an external input event} |
event $\T<type>$ $\T<id>$; | \CC{19.5em}{declares an internal event}       |
var   $\T<type>$ $\T<id>$; | \CC{19.5em}{declares a variable}              |

|\C{Event handling}|
$\T<id>$ = await $\T<id>$; | \CC{19.5em}{awaits an event and assigns the received value} |
emit $\T<id>$($\T<exp>$); | \CC{19.5em}{emits an event passing a value} |

|\C{Control flow}|
$\T<stmt>$ ; $\T<stmt>$ | \CC{10em}{sequence}     |
if $\T<exp>$ then $\T<stmts>$ else $\T<stmts>$ end | \CC{10em}{conditional}  |
loop do $\T<stmts>$ end | \CC{10em}{repetition}   |
every $\T<id>$ in $\T<id>$ do $\T<stmts>$ end | \CC{10em}{event iteration}   |
finalize [$\T<stmts>$] with $\T<stmts>$ end | \CC{10em}{finalization} |

|\C{Logical parallelism}|
par/or  do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{aborts when any side ends}      |
par/and do $\T<stmts>$ with $\T<stmts>$ end | \CC{13em}{terminates when all sides end} |

|\C{Assignment \& Integration with C}|
$\T<id>$ = $\T<exp>$; | \CC{19.5em}{assigns a value to a variable}              |
_$\T<id>$($\T<exps>$) | \CC{19.5em}{calls a C function (id starts with `\_'\,)} |
\end{lstlisting}
\egroup

Listing~\ref{lst.ceu} shows a complete example in \CEU that toggles a LED
whenever a radio packet is received, terminating on a button press always
with the LED off.
%
The program first declares the \code{BUTTON} and \code{RADIO\_RECV} as
input events (ln.~1--2).
Then, it uses a \code{par/or} composition to run two activities in parallel:
a single-statement trail that waits for a button press before terminating
(ln.~4), and an endless loop that toggles the LED on and off on radio receives
(ln.~9--14).
The \code{finalize} clause (ln.~6--8) ensures that, no matter how its enclosing
trail terminates, the LED will be unconditionally turned off~(ln.~7).

@@

@@

\begingroup
\begin{figure}[h]
\centering
\begin{tikzpicture}[x=.9em,y=.9em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\scriptsize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Real World)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\textbf{A}};
  \node[evt](Bevt) at (B) {\textbf{B}};
  \node[evt](Cevt) at (C) {\textbf{C}};
  \node[evt](Devt) at (D) {\textbf{D}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(\CEU)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

@@

\begingroup
\begin{figure}[t]
\begin{minipage}[h]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\small{\ax Accesses to \code{x} are never\\\phantom{\ax}\relax concurrent.}
\end{minipage}%
%
\begin{minipage}[h]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

\end{lstlisting}
\small{\bx Accesses to \code{y} are concurrent\\
\phantom{\bx}\relax but still deterministic.}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{Shared-memory concurrency in \CEU:
\ax is safe because the trails access \code{x} atomically in different
reactions;
\bx is unsafe because both trails access \code{y} in the same reaction}
\label{lst.shared}
\end{figure}
\endgroup

@@

\begin{figure}[h]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\footnotesize]
par/or do
   var _msg_t msg;
   <...> // prepare msg
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
//
\end{lstlisting}
\small{\ax\!Local resource finalization\rlap{.}}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=2em,basicstyle=\ttfamily\footnotesize]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\small{\quad\bx\!External resource finalization\rlap{.}}
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
\label{lst.fin.ceu}
\end{figure}

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}.
%
%\begin{itemize}
 If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
 If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
%\end{itemize}

\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}
